MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-08-08 latex/paper/sections/04-architecture_history.tex
more packing
start talking about ap
more ap

\textbf{Homogeneous FPGA:} \quad 
The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}. 
The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
\\

{
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
    \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
    \label{fig:cpld}
    % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
}
{
    \centering
    \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
    \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
    \label{fig:fpga}
    % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
}

@@

{
    \centering
    \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
    \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
    \label{fig:fpga}
    % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
}
\vspace{0.5cm}

\textbf{Heterogenous FPGA:} \quad 
This brings us to modern day FPGA architectures. 
To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}. 
This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

>>>>

\textbf{Homogeneous FPGA:} \quad 
The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:homogen_fpga}. 
The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
\\

{
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
    \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
    \label{fig:cpld}
    % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
}
{
    \centering
    \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
    \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
    \label{fig:homogen_fpga}
    % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
}

@@

{
    \centering
    \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
    \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
    \label{fig:heterogen_fpga}
    % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
}
\vspace{0.25cm}

\textbf{Heterogenous FPGA:} \quad 
This brings us to modern day FPGA architectures. 
To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:heterogen_fpga}. 
This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
Major vendors such as Xilinx (AMD) and Altera (Intel) now employ heterogeneous island-style architectures in their devices. 
As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

### 2025-08-08 latex/paper/sections/09-packing.tex
more reorg
more ap

Listing \ref{lst:siteinst} shows how to use the \texttt{SiteInst} object and how it ties \texttt{Site}s, \texttt{Cell}s, \texttt{BEL}s, and intra-Site \texttt{Net}s together. 
The super-specifics of the packing process via RapidWright is too involved for the scope of this paper, but just to give an impression of how the \texttt{design} classes are used, Listing \ref{lst:single_carry_chains}, \ref{lst:carry_chains}, \ref{lst:carry_nets}, and \ref{lst:ff_nets} together show how \texttt{CarryCellGroup}s are packed into \texttt{SLICEL} \texttt{Sites} in our working code.

Packing the other \texttt{Cell} structures into \texttt{Site}s follow a similar process. 
For each cell cluster, we create a \texttt{SiteInst} and populate its \texttt{BEL}s with the corresponding \texttt{Cell}s in the cluster and assign it a physical \texttt{Site} on the \texttt{device}.
We further consolidate these \texttt{SiteInst}s into clusters or lists of \texttt{SiteInst}s for multi-Site structures like \texttt{CarryCellGroup} chains and DSP cascades. 
For each \texttt{SiteInst}, we assign it a physical \texttt{Site} on the \texttt{device} and route its internal intra-Site nets. 
The end result is a \texttt{PackedDesign} object which is a group of data structures solely around \texttt{SiteInst}s like shown in figure \ref{fig:substages}.
At this point we no longer need to worry about individual \texttt{BEL}s, \texttt{Cell}s, or intra-Site \texttt{Nets} as they have now been packaged and taped-off into their respective \texttt{SiteInst}s. 
These \texttt{SiteInst} become the new atomic level components in the following Placement stage. 
This process is analogous to the containerization of loose goods in ocean freight forwarding.

@@

\newpage
\begin{lstlisting}[language=java, caption={Packing \texttt{CarryCellGroup}s into \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:carry_chains}]
// Pack all CarryCellGroup chains
private List<List<SiteInst>> packCarryChains(List<List<CarryCellGroup>> EDIFCarryChains)
        throws IOException {
    List<List<SiteInst>> siteInstChains = new ArrayList<>();
    writer.write("\n\nPacking carry chains... (" + EDIFCarryChains.size() + ")");
    for (List<CarryCellGroup> edifChain : EDIFCarryChains) {
        List<SiteInst> siteInstChain = new ArrayList<>();
        writer.write("\n\t\tChain Size: (" + edifChain.size() + "), Chain Anchor: "
                + edifChain.get(0).carry().getFullHierarchicalInstName());
        Site anchorSite = selectCarryAnchorSite(edifChain.size());
        SiteTypeEnum selectedSiteType = anchorSite.getSiteTypeEnum();
        for (int i = 0; i < edifChain.size(); i++) {
            Site site = (i == 0) ? anchorSite
                    : device.getSite("SLICE_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
            SiteInst si = new SiteInst(edifChain.get(i).carry().getFullHierarchicalInstName(), design,
                    selectedSiteType,
                    site);
            packCarrySite(edifChain.get(i), si);
            if (i == 0) { // additional routing logic for anchor site
                Net CINNet = si.getNetFromSiteWire("CIN");
                CINNet.removePin(si.getSitePinInst("CIN"));
                si.addSitePIP(si.getSitePIP("PRECYINIT", "0"));
            }
            occupiedSites.get(selectedSiteType).add(site);
            availableSites.get(selectedSiteType).remove(site);
            siteInstChain.add(si);
        }
        siteInstChains.add(siteInstChain);
    } // end for (List<EDIFCellInst> chain : EDIFCarryChains)
    return siteInstChains;
} // end packCarryChains()
\end{lstlisting}

>>>>

Listing \ref{lst:siteinst} shows how to use the \texttt{SiteInst} object and how it ties \texttt{Site}s, \texttt{Cell}s, \texttt{BEL}s, and intra-Site \texttt{Net}s together. 
The super-specifics of the packing process via RapidWright is too involved for the scope of this paper, but just to give an impression of how the \texttt{design} classes are used, Listing \ref{lst:single_carry_chains}, \ref{lst:carry_cell_group_chains}, \ref{lst:carry_nets}, and \ref{lst:ff_nets} together show how \texttt{CarryCellGroup}s are packed into \texttt{SLICEL} \texttt{Sites} in our working code.

Packing the other \texttt{Cell} structures into \texttt{Site}s follow a similar process. 
For each cell cluster, we create a \texttt{SiteInst} and populate its \texttt{BEL}s with the corresponding \texttt{Cell}s in the cluster and assign it a physical \texttt{Site} on the \texttt{device}.
We further consolidate these \texttt{SiteInst}s into clusters or lists of \texttt{SiteInst}s for multi-Site structures like \texttt{CarryCellGroup} chains and DSP cascades. 
For each \texttt{SiteInst}, we assign it a physical \texttt{Site} on the \texttt{device} and route its internal intra-Site nets. 
The end result is a \texttt{PackedDesign} object which is a group of data structures solely around \texttt{SiteInst}s like shown in figure \ref{fig:substages}.
The process is analogous to containerization of loose goods in ocean freight forwarding. 
At this point we no longer need to worry about individual \texttt{BEL}s, \texttt{Cell}s, or intra-Site \texttt{Nets} as they have now been packaged and taped-off into their respective \texttt{SiteInst}s. 
These \texttt{SiteInst} become the new atomic level components in the following Placement stage.

@@

\newpage
\begin{lstlisting}[language=java, caption={Packing \texttt{CarryCellGroup}s into \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:carry_cell_group_chains}]
// Pack all CarryCellGroup chains
private List<List<SiteInst>> packCarryChains(List<List<CarryCellGroup>> EDIFCarryChains)
        throws IOException {
    List<List<SiteInst>> siteInstChains = new ArrayList<>();
    writer.write("\n\nPacking carry chains... (" + EDIFCarryChains.size() + ")");
    for (List<CarryCellGroup> edifChain : EDIFCarryChains) {
        List<SiteInst> siteInstChain = new ArrayList<>();
        writer.write("\n\t\tChain Size: (" + edifChain.size() + "), Chain Anchor: "
                + edifChain.get(0).carry().getFullHierarchicalInstName());
        Site anchorSite = selectCarryAnchorSite(edifChain.size());
        SiteTypeEnum selectedSiteType = anchorSite.getSiteTypeEnum();
        for (int i = 0; i < edifChain.size(); i++) {
            Site site = (i == 0) ? anchorSite
                    : device.getSite("SLICE_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
            SiteInst si = new SiteInst(edifChain.get(i).carry().getFullHierarchicalInstName(), design,
                    selectedSiteType,
                    site);
            packCarrySite(edifChain.get(i), si);
            if (i == 0) { // additional routing logic for anchor site
                Net CINNet = si.getNetFromSiteWire("CIN");
                CINNet.removePin(si.getSitePinInst("CIN"));
                si.addSitePIP(si.getSitePIP("PRECYINIT", "0"));
            }
            occupiedSites.get(selectedSiteType).add(site);
            availableSites.get(selectedSiteType).remove(site);
            siteInstChain.add(si);
        }
        siteInstChains.add(siteInstChain);
    } // end for (List<EDIFCellInst> chain : EDIFCarryChains)
    return siteInstChains;
} // end packCarryChains()
\end{lstlisting}

### 2025-08-08 latex/paper/sections/11-placement_results.tex
more ap

Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan), which poses a computational challenge for the legalization process in directed placement.

>>>>

Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan), which poses a computational challenge for the legalization process in directed and analytical placement.

### 2025-08-08 latex/paper/sections/12-future_work.tex
more ap

\subsection{Improve Robustness}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area. 
Enabling \texttt{SLICEL}-\texttt{SLICEM} homogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.

@@

\subsection{Force-Directed and Analytical Placement}
As mentioned previously, SA has largely been abandoned in favor of AP in SOTA placers, owing to SA's poor scalability and runtime.
Our placer follows a basic prepack-pack-place flow, where the final placement stage is driven by simulated annealing. 
In future work, we can retrofit this flow by replacing SA with AP in the final placement stage while fully reusing the prepacker and packer in its current state. 
We can still treat \texttt{SiteInst}s as the atomic objects to be placed and use analytical solvers to minimize the total HPWL of the nets between them.

The following papers can serve as foundational references for building an AP placer or insight into the SOTA placers.
\begin{itemize}
    \item \emph{Analytical minimization of half-permieter wirelength} - \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
    \item \emph{SimPL: an algorithm for placing VLSI circuits} - \textbf{Kim et al. (2013)} \cite{SimPL}.
    \item \emph{Analytical placement for heterogeneous FPGAs} - \textbf{Gort et al. (2012)} \cite{AP_2012}.
    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL-SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} - \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macros Groups and Fence Regions} - \textbf{Jing et al. (2024)} \cite{OpenPARF}.
\end{itemize}

@@

\begin{equation}
    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i, j} w_{i, j} ( |x_i - x_j| + |y_i - y_j| )
\end{equation}

>>>>

\subsection{Improve Robustness}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} heterogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area. 
Enabling \texttt{SLICEL}-\texttt{SLICEM} heterogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.

@@

\subsection{Force-Directed and Analytical Placement}
As discussed earlier, simulated annealing (SA) has largely been phased out in state-of-the-art (SOTA) placers due to its poor scalability and long runtimes. 
Our current placer follows a straightforward \emph{prepack–pack–place} flow, with the final placement stage driven by SA. 
In future work, this stage can be retrofitted to use analytical placement (AP) while reusing the prepacker and packer. 
Here, \texttt{SiteInst}s remain the atomic placement objects, and analytical solvers can be applied to minimize the total half-perimeter wirelength (HPWL) of the connecting nets.

The following works provide a foundation for understanding AP, ranging from early formulations to modern SOTA methods:
\begin{itemize}
    \item \emph{Analytical minimization of half-perimeter wirelength} — \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
    \item \emph{Analytical placement for heterogeneous FPGAs} — \textbf{Gort et al. (2012)} \cite{AP_2012}.
    \item \emph{SimPL: An algorithm for placing VLSI circuits} — \textbf{Kim et al. (2013)} \cite{SimPL}.
    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL–SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} — \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macro Groups and Fence Regions} — \textbf{Jing et al. (2024)} \cite{OpenPARF}.
\end{itemize}

As noted in Section~\ref{sec:placement}, replacing SA with AP requires splitting the placement stage into two substages:  
(1) \emph{global placement}, which determines continuous target positions using analytical optimization, and  
(2) \emph{detailed placement} (legalization), which snaps objects to valid \texttt{Site}s while adhering to hardware constraints.  
Global placement can be performed using any number of off-the-shelf (OTS) analytical solvers.

Unlike SA, which relies primarily on heuristics and probabilistic moves, AP begins with a formal mathematical formulation that is compatible with analytic optimization. 
As introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a \emph{hypergraph}
\[
    G_{H} = (V_{H}, E_{H}),
\]
where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
In the FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s interconnecting them.

We can then reduce this hypergraph into a graph by converting all multi-pin hyperedges into a set of 2-pin edges.

Most AP approaches then minimize the weighted sum of half perimeter wirelengths (HPWL) of these 2-pin edges, with some approaches using the sum of linear distances \ref{equ:Manhattan} and others using the sum of squared distances \ref{equ:Euclidian}.

\begin{equation}
    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
    \label{equ:Manhattan}
\end{equation}

\begin{equation}
    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
    \label{equ:Euclidian}
\end{equation}

% use these formulations
% HPWL_AP_2000
% HPWL_AP_2012

@@
